Introduction to Causal Inference Models

### 2017-07-12 msm_introduction.Rmd
moved my scripts to this general repo about methods.

2. Marginal structural models 
- Sometimes called inverse-probability-of-treatment weighting or IPTW

@@

## A Mostly Accurate History of Causal Models

A long time ago (1920s), Jerzy Neyman proposed a class of statistical models that would latter be referred to as counter factual/potential outcomes models for randomized trials. Donald Rubin extended this framework to observational data and coined the term potential outcomes in maybe the 1970s?. Judea Pearl and Jamie Robins also did some stuff in the 80's. Jamie Robins introduced the g-formula and IPTW which were designed to estimate the marginal effect of a binary treatment that varied over time and was confounded by a time-varying confounder. Jamie Robins and a lot of his students have made some tweaks to the original framework which pretty much brings us to today.

Anyways, a major process/step to most of these early models is the concept of the counter factual (which we'll cover in depth later in this presentation).

The general concept of the counter factual is it takes our "what if?" questions from implicit to explicit in our data.

@@

*3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a = 1 and a = 0*

@@

*2x2 Table*
```{r 2x2}
xtabs(~A + Y, sim_data)
```

*Risk Ratio of E(Y|A)*

```{r ratio}
# standard ratio model
ratio_model <- glm(Y~A, sim_data, family = "binomial"(link="logit"))
# ratio estimate
exp(ratio_model$coefficients[2])
```
The treated group is 40% less likely to develop the outcome compared to the untreated group.

## Updating the Predicted Outcomes Under a = 1

First we duplicate our original dataframe and then set everyone's treatment to 1.

```{r create a1 exposure dataframe}
# set everyone to exposed from our original dataframe
sim_data_a1 <- sim_data %>% mutate(A = 1)
# view first 6 records
head(sim_data_a1)
```

Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the dataframe where everyone receives treatment E(Y|a=1).

```{r pr a1}
# updated predicted probability of outcome 
pr_a1 <- predict(ratio_model, newdata = sim_data_a1, type = "response")
# average probability where everyone is exposed
mean(pr_a1)
```

## Updating the Predicted Outcomes Under a = 0

We repeat the same step under the scenario where no one is treated.

```{r create a1 exposure dataframe}
# set everyone to exposed from our original dataframe
sim_data_a0 <- sim_data %>% mutate(A = 0)
# view first 6 records
head(sim_data_a0)
```

Updating our predicted probability under no treatment scenario E(Y|a=0).

```{r pr a0}
# updated predicted probability of outcome 
pr_a0 <- predict(ratio_model, newdata = sim_data_a0, type = "response")
# average probability where no one is exposed
mean(pr_a0)
```

## Estimate the Counterfactual Parameter of Interest

Now that we have a mean value for Y where everyone is treated and no one is treated, we can estimate our risk ratio.

```{r g forumla ratio}
mean(pr_a1)/mean(pr_a0)
```

Let's compare to our estimate using logistic regression again.

```{r standard est result}
exp(ratio_model$coefficients[2])
```

We get pretty much the same result. That was a lot of steps to get the same answer.

>>>>

2. Marginal structural models 
- Sometimes called inverse-probability-of-treatment weighting (IPTW)

@@

## A Brief History of Causal Models

A long time ago (1920s), in a country far far away (Poland), Jerzy Neyman proposed a class of statistical models that would latter be referred to as counter factual/potential outcomes models for randomized trials. Donald Rubin extended this framework to observational data and coined the term potential outcomes in maybe the 1970s?. Judea Pearl and Jamie Robins also did some stuff in the 80's. Jamie Robins introduced the g-formula and IPTW which were designed to estimate the marginal effect of a binary treatment that varied over time and was confounded by a time-varying confounder. Jamie Robins and a lot of his students have made some tweaks to the original framework which pretty much brings us to today.

## The Counterfactual

- A major process/step to most of these early models is the concept of the counter factual (which we'll cover in depth later in this presentation).

- The general concept of the counter factual is it takes our "what if?" questions from implicit to explicit in our data.

@@

*3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a=1 and a=0*

@@

*2x2 Table*
```{r 2x2}
xtabs(~A + Y, sim_data)
```
*Risk Ratio of E(Y|A)*
```{r ratio}
# standard ratio model
ratio_model <- glm(Y~A, sim_data, family = "binomial"(link="logit"))
# ratio estimate
exp(ratio_model$coefficients[2])
```
The treated group is 40% less likely to develop the outcome compared to the untreated group.

## Updating the Predicted Outcomes Under a = 1 
First we duplicate our original dataframe and then set everyone's treatment to 1.
```{r create a1 exposure dataframe}
# set everyone to exposed from our original dataframe
sim_data_a1 <- sim_data %>% mutate(A = 1)
# view first 6 records
head(sim_data_a1)
```
Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the dataframe where everyone receives treatment E(Y|a=1).
```{r pr a1}
# updated predicted probability of outcome 
pr_a1 <- predict(ratio_model, newdata = sim_data_a1, type = "response")
# average probability where everyone is exposed
mean(pr_a1)
```

## Updating the Predicted Outcomes Under a = 0
We repeat the same step under the scenario where no one is treated.
```{r create a0 exposure dataframe}
# set everyone to exposed from our original dataframe
sim_data_a0 <- sim_data %>% mutate(A = 0)
# view first 6 records
head(sim_data_a0)
```
Updating our predicted probability under no treatment scenario E(Y|a=0).
```{r pr a0}
# updated predicted probability of outcome 
pr_a0 <- predict(ratio_model, newdata = sim_data_a0, type = "response")
# average probability where no one is exposed
mean(pr_a0)
```

## Estimate the Counterfactual Parameter of Interest
Now that we have a mean value for Y where everyone is treated and no one is treated, we can estimate our risk ratio.
```{r g forumla ratio}
# g formula ratio
mean(pr_a1)/mean(pr_a0)
```
Let's compare to our estimate using logistic regression again.
```{r standard est result}
# odds ratio (approximates risk ratio due to rare disease assumption)
exp(ratio_model$coefficients[2])
```

We get pretty much the same result using the g-formula approach to a standard regression.

## Counterfactual Estimate using IPTW
